---
title: "Getting Started with SWORD"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with SWORD}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  fig.width = 6,
  fig.height = 4,
  out.width = "95%"
)
```

## Overview

**SWORD** (**S**upport vector machine **W**eighted **O**blique **R**andom
**D**ecision forests) fits random forests of oblique regression trees.
Each internal node uses a **Weighted-SVM** classifier with a linear kernel
to define the oblique splitting hyperplane — a sparse linear combination of
predictors — rather than a single axis-aligned threshold.

The base learner is **TORS** (**T**ree **O**blique for **R**egression with
weighted **S**VM), a single oblique regression tree introduced in
Carta & Frigau (2025).  SWORD embeds TORS within a bootstrap aggregation
framework, adding predictor subspacing and optional randomisation of the
number of features per split to increase tree diversity.

The package provides:

* `TORS()` — fit a single TORS oblique regression tree.
* `SWORD()` — fit a SWORD forest of TORS trees with bootstrap sampling and
  out-of-bag (OOB) error estimation.
* `predict()` — S3 method for both objects, with automatic encoding of factor
  and character predictors.
* `VI_SWORD()` — Oblique Impurity-Weighted Variable Importance (OIW-VI):
  deviance reduction at each node, distributed among predictors proportionally
  to their normalised absolute Weighted-SVM coefficients.
* Plot helpers — OOB convergence, importance barplot, predicted-vs-observed
  scatter, and partial dependence plots.

---

## Data

We use the **Boston** housing dataset from the `MASS` package (506
observations, 13 numeric predictors, response `medv` = median house value).
The data are split 80/20 into training and test sets.

```{r data}
if (!requireNamespace("MASS", quietly = TRUE))
  stop("Please install the MASS package: install.packages('MASS')")

library(SWORD)
data(Boston, package = "MASS")

set.seed(42)
n     <- nrow(Boston)
train <- sample(n, floor(0.8 * n))
test  <- setdiff(seq_len(n), train)

X_train <- Boston[train, setdiff(names(Boston), "medv")]
y_train <- Boston$medv[train]
X_test  <- Boston[test,  setdiff(names(Boston), "medv")]
y_test  <- Boston$medv[test]

cat("Training set:", nrow(X_train), "obs |",
    "Test set:", nrow(X_test), "obs\n")
```

---

## Fitting a single TORS tree

`TORS()` grows one SVM oblique rotation tree.  Key tuning parameters:

| Parameter | Meaning | Default |
|---|---|---|
| `nmin` | minimum node size | 5 |
| `cp` | complexity threshold for splitting | 0.01 |
| `n_topCor` | top-correlated features fed to the SVM at each node | 2 |
| `relation` | feature selection criterion: `"Pearson"` or `"MI"` | `"Pearson"` |
| `Weight_Scheme` | observation weighting: `"scale"` or `"robust"` | `"scale"` |
| `cost_C` | SVM cost C | 1 |

```{r tors-fit}
set.seed(1)
tree <- TORS(X_train, y_train, nmin = 15, cp = 0.02)
print(tree)
```

### Tree structure (text)

`plot_tors_text()` prints the tree as an indented list of nodes.  Each
internal node shows the oblique hyperplane equation (top-`k` largest
coefficients); each leaf shows the predicted mean.

```{r tors-text, fig.height=5}
plot_tors_text(tree, top_k = 3)
```

### Tree structure (graph)

`plot()` on a `tors_flat` object draws a base-R layout of the tree (node size
proportional to the number of observations).  If the **visNetwork** package is
installed it renders an interactive zoomable widget instead.

```{r tors-plot}
plot(tree)
```

### Effect of `cp` on tree size

`cp` is a complexity threshold: a candidate split is accepted only if the
relative deviance reduction exceeds `cp`.  Larger values lead to earlier
stopping and shallower trees; `cp = 0` lets the tree grow until every node
reaches `nmin`.

```{r tors-cp}
set.seed(1)
t_deep     <- TORS(X_train, y_train, nmin = 10, cp = 0.000)
t_moderate <- TORS(X_train, y_train, nmin = 10, cp = 0.010)
t_shallow  <- TORS(X_train, y_train, nmin = 10, cp = 0.050)

cp_tbl <- data.frame(
  cp        = c(0.000, 0.010, 0.050),
  n_nodes   = c(t_deep$n_nodes,     t_moderate$n_nodes,     t_shallow$n_nodes),
  n_leaves  = c(sum(t_deep$is_leaf), sum(t_moderate$is_leaf), sum(t_shallow$is_leaf)),
  max_depth = c(max(t_deep$depth),   max(t_moderate$depth),   max(t_shallow$depth))
)
print(cp_tbl, row.names = FALSE)
```

With `cp = 0` the tree grows until every node reaches `nmin`, producing the
deepest structure; increasing `cp` results in more aggressive pruning and
shallower trees.

### Effect of `n_topCor` on split richness

`n_topCor` controls how many top-correlated features are passed to the
Weighted-SVM at each node.  Because TORS splits are always oblique (linear
combinations of predictors), the minimum meaningful value is 2; larger values
allow the SVM to combine more predictors into denser oblique hyperplanes.

```{r tors-ntopcor}
set.seed(1)
t_nc2 <- TORS(X_train, y_train, nmin = 15, cp = 0.02, n_topCor = 2)
t_nc4 <- TORS(X_train, y_train, nmin = 15, cp = 0.02, n_topCor = 4)

cat("--- n_topCor = 2 (default, sparse oblique) ---\n")
plot_tors_text(t_nc2, top_k = 2)
cat("\n--- n_topCor = 4 (richer oblique splits) ---\n")
plot_tors_text(t_nc4, top_k = 4)
```

With `n_topCor = 2` each split combines two predictors; with `n_topCor = 4`
the Weighted-SVM can involve up to four predictors per split, producing denser
oblique hyperplanes at the cost of higher per-node fitting time.

---

## Fitting a SWORD forest

`SWORD()` fits a random forest of TORS trees with bootstrap sampling.
Setting `OOB = TRUE` (the default) accumulates out-of-bag predictions at each
iteration, which lets you monitor convergence without a separate validation set.

```{r sword-fit}
set.seed(42)
forest <- SWORD(X_train, y_train,
                m       = 50,
                nmin    = 10,
                cp      = 0.01,
                OOB     = TRUE,
                verbose = FALSE)
print(forest)
```

---

## OOB diagnostics

### Convergence curve

`plot(forest, which = "oob")` plots the OOB MSE as a function of the number
of trees.  The dashed line marks the minimum; check that it has levelled off
before using the forest for inference.

```{r oob-convergence}
plot(forest, which = "oob")
```

### OOB predicted vs. observed

`plot_oob_fit()` plots the out-of-bag predictions against the true training
responses.  Points on the identity line indicate perfect fit; the annotation
reports RMSE, MAE, and R\eqn{^2}.

```{r oob-scatter}
plot_oob_fit(forest, y_train)
```

---

## Prediction on the test set

```{r predict}
y_hat <- predict(forest, X_test)

rmse <- sqrt(mean((y_hat - y_test)^2))
mae  <- mean(abs(y_hat - y_test))
r2   <- cor(y_hat, y_test)^2

cat(sprintf("Test RMSE : %.3f\n", rmse))
cat(sprintf("Test MAE  : %.3f\n", mae))
cat(sprintf("Test R^2  : %.3f\n", r2))
```

The `predict()` method also accepts a `data.frame` that contains extra columns
(e.g. the response), so you can pass the original data frame directly:

```{r predict-full-df}
y_hat2 <- predict(forest, Boston[test, ])  # medv column is silently dropped
all.equal(y_hat, y_hat2)
```

---

## Variable importance

`VI_SWORD()` computes the **OIW-VI** (Oblique Impurity-Weighted Variable
Importance): at each internal node, the deviance reduction is distributed
among the predictors involved in the oblique split proportionally to their
normalised absolute Weighted-SVM coefficients; these scores are then averaged
across all trees.  `plot_vi_sword()` draws a horizontal barplot sorted by
importance.

```{r vi}
vi <- VI_SWORD(forest)
print(round(vi, 2))
```

```{r vi-plot}
plot_vi_sword(forest, top_n = 13)
```

---

## Partial dependence plots

`pdp_sword()` computes the marginal effect of a single predictor by fixing it
to a grid of values and averaging forest predictions over the full training set
(Friedman 2001).  A rug of observed values is added by default.

```{r pdp-2var, fig.height=4, fig.width=10, out.width="100%"}
old_par <- par(mfrow = c(1, 2), mar = c(4, 4, 3, 1))
pdp_sword(forest, X_train, "lstat", main = "PDP: lstat")
pdp_sword(forest, X_train, "rm",    main = "PDP: rm")
par(old_par)
```

Top-4 predictors in a 2x2 grid:

```{r pdp-grid, fig.height=7, fig.width=10, out.width="100%"}
top4 <- names(vi)[1:4]
old_par <- par(mfrow = c(2, 2), mar = c(4, 4, 3, 1))
for (v in top4)
  pdp_sword(forest, X_train, v, main = paste("PDP:", v))
par(old_par)
```

---

## Parameter sensitivity in SWORD

Four parameters most likely to affect predictive accuracy are `relation`,
`Weight_Scheme`, `rf_var`, and `rand_ntopcor`.  In this section we fit
compact forests (`m = 20`, `nmin = 10`, `cp = 0.01`) varying each parameter
in turn and compare test-set RMSE on the same Boston hold-out.

```{r params-helper, include=FALSE}
rmse_f <- function(y, yhat) sqrt(mean((y - yhat)^2))
p      <- ncol(X_train)
```

### Feature selection criterion: `relation`

`relation` controls how the top-correlated features are identified at each
node: `"Pearson"` uses the absolute Pearson correlation with the node
residuals; `"MI"` uses mutual information, which can capture non-linear
associations between each predictor and the response.

```{r params-relation}
set.seed(42)
f_pearson <- SWORD(X_train, y_train, m = 20, nmin = 10, cp = 0.01,
                   relation = "Pearson", verbose = FALSE)
set.seed(42)
f_mi      <- SWORD(X_train, y_train, m = 20, nmin = 10, cp = 0.01,
                   relation = "MI",      verbose = FALSE)

cat(sprintf("Pearson  — test RMSE: %.3f\n", rmse_f(y_test, predict(f_pearson, X_test))))
cat(sprintf("MI       — test RMSE: %.3f\n", rmse_f(y_test, predict(f_mi,      X_test))))
```

Both criteria are competitive on continuous predictors; `"MI"` may offer an
advantage when predictor–response associations are strongly non-linear.

### Observation weighting: `Weight_Scheme`

`Weight_Scheme` determines how observations are weighted when fitting the
Weighted-SVM at each node.  Both schemes assign higher weight to observations
whose response is far from the local centre, guiding the SVM to concentrate
on the most informative thresholds.  The two formulas differ in how the
centre and spread are estimated:

\[
w_i^{\text{scale}}  = \frac{|y_i - \bar{y}|}{s_y}, \qquad
w_i^{\text{robust}} = \frac{|y_i - \text{median}(y)|}{\text{MAD}(y)}
\]

where \(s_y\) is the standard deviation and
\(\text{MAD}(y) = \text{median}(|y_i - \text{median}(y)|)\).
Because `"robust"` centres on the median and scales by the MAD rather than
the mean and standard deviation, it is less influenced by extreme values in
the response.

```{r params-weight}
set.seed(42)
f_scale  <- SWORD(X_train, y_train, m = 20, nmin = 10, cp = 0.01,
                  Weight_Scheme = "scale",  verbose = FALSE)
set.seed(42)
f_robust <- SWORD(X_train, y_train, m = 20, nmin = 10, cp = 0.01,
                  Weight_Scheme = "robust", verbose = FALSE)

cat(sprintf("scale    — test RMSE: %.3f\n", rmse_f(y_test, predict(f_scale,  X_test))))
cat(sprintf("robust   — test RMSE: %.3f\n", rmse_f(y_test, predict(f_robust, X_test))))
```

`"robust"` is recommended when the training response contains outliers or
extreme values that would inflate \(s_y\) and distort the `"scale"` weights.

### Feature subsampling per split: `rf_var`

`rf_var` sets how many predictors are randomly sub-sampled at each node
before the top-`n_topCor` correlated ones are selected for the SVM.
Reducing `rf_var` increases tree decorrelation — analogously to `mtry` in
standard random forests.  Three common choices with `p = 13` predictors:

```{r params-rfvar}
set.seed(42)
f_full <- SWORD(X_train, y_train, m = 20, nmin = 10, cp = 0.01,
                rf_var = p,               verbose = FALSE)  # all predictors
set.seed(42)
f_half <- SWORD(X_train, y_train, m = 20, nmin = 10, cp = 0.01,
                rf_var = floor(p / 2),    verbose = FALSE)  # half
set.seed(42)
f_sqrt <- SWORD(X_train, y_train, m = 20, nmin = 10, cp = 0.01,
                rf_var = floor(sqrt(p)),  verbose = FALSE)  # sqrt rule

cat(sprintf("rf_var = %2d (all)        — test RMSE: %.3f\n",
            p,               rmse_f(y_test, predict(f_full, X_test))))
cat(sprintf("rf_var = %2d (floor p/2)  — test RMSE: %.3f\n",
            floor(p / 2),   rmse_f(y_test, predict(f_half, X_test))))
cat(sprintf("rf_var = %2d (floor sqrt) — test RMSE: %.3f\n",
            floor(sqrt(p)), rmse_f(y_test, predict(f_sqrt, X_test))))
```

Unlike axis-aligned forests, where reducing `mtry` decorrelates trees and
often lowers test error, in SWORD a **larger candidate pool consistently
improves accuracy**: allowing the Weighted-SVM to screen more predictors at
each node leads to better oblique splits without sacrificing tree diversity
(which is already provided by bootstrap sampling and the random-\eqn{\gamma}
mechanism).  Restricting `rf_var` limits the quality of the oblique
hyperplane and in practice tends to worsen RMSE.  The default
(`rf_var = p`, all predictors) is therefore recommended.

### Randomised top-feature cardinality: `rand_ntopcor`

When `rand_ntopcor = TRUE` (the SWORD default) the number of top-correlated
features passed to the SVM at each node is itself drawn at random from
\eqn{\{2, \ldots, \texttt{rf\_var}\}}, introducing an extra source of tree
diversity beyond bootstrap sampling.  Setting it to `FALSE` fixes that count
at `n_topCor` for every split.

```{r params-rand}
set.seed(42)
f_rnd_t <- SWORD(X_train, y_train, m = 20, nmin = 10, cp = 0.01,
                 rand_ntopcor = TRUE,  verbose = FALSE)
set.seed(42)
f_rnd_f <- SWORD(X_train, y_train, m = 20, nmin = 10, cp = 0.01,
                 rand_ntopcor = FALSE, verbose = FALSE)

cat(sprintf("rand_ntopcor = TRUE  — test RMSE: %.3f\n",
            rmse_f(y_test, predict(f_rnd_t, X_test))))
cat(sprintf("rand_ntopcor = FALSE — test RMSE: %.3f\n",
            rmse_f(y_test, predict(f_rnd_f, X_test))))
```

Randomisation usually helps when `m` is large enough; for very small forests
it can occasionally hurt by introducing excess variance.

### Summary comparison

```{r params-table, echo=FALSE}
tbl <- data.frame(
  Parameter = c(
    "relation", "relation",
    "Weight_Scheme", "Weight_Scheme",
    "rf_var", "rf_var", "rf_var",
    "rand_ntopcor", "rand_ntopcor"
  ),
  Value = c(
    "Pearson (default)", "MI",
    "scale (default)", "robust",
    paste0("p = ", p, " — all (default)"),
    paste0("floor(p/2) = ", floor(p / 2)),
    paste0("floor(sqrt(p)) = ", floor(sqrt(p))),
    "TRUE (default)", "FALSE"
  ),
  Test_RMSE = round(c(
    rmse_f(y_test, predict(f_pearson, X_test)),
    rmse_f(y_test, predict(f_mi,      X_test)),
    rmse_f(y_test, predict(f_scale,   X_test)),
    rmse_f(y_test, predict(f_robust,  X_test)),
    rmse_f(y_test, predict(f_full,    X_test)),
    rmse_f(y_test, predict(f_half,    X_test)),
    rmse_f(y_test, predict(f_sqrt,    X_test)),
    rmse_f(y_test, predict(f_rnd_t,   X_test)),
    rmse_f(y_test, predict(f_rnd_f,   X_test))
  ), 3)
)
print(tbl, row.names = FALSE)
```

---

## Categorical and character predictors

SWORD handles factor and character columns automatically: they are
one-hot-encoded internally, new factor levels in `predict()` are mapped to
all-zero dummies, and `VI_SWORD()` sums dummy-column importances back to the
original variable name.

```{r categorical, eval=FALSE}
data(diamonds, package = "ggplot2")
set.seed(7)
idx   <- sample(nrow(diamonds), 2000)
X_dia <- diamonds[idx, c("carat","cut","color","clarity",
                          "depth","table","x","y","z")]
y_dia <- diamonds$price[idx]

forest_dia <- SWORD(X_dia, y_dia, m = 50, OOB = TRUE, verbose = FALSE)
print(forest_dia)

# VI grouped by original variable (not by dummy column)
VI_SWORD(forest_dia)

# PDP for an ordered factor
pdp_sword(forest_dia, X_dia, "cut")
```

---

## Formula interface

Both `TORS()` and `SWORD()` accept an R formula with a `data` argument:

```{r formula, eval=FALSE}
df <- cbind(X_train, medv = y_train)

tree2   <- TORS(medv ~ ., data = df, nmin = 15, cp = 0.02)
forest2 <- SWORD(medv ~ ., data = df, m = 50, verbose = FALSE)
predict(forest2, Boston[test, ])
```

---

## Summary

| Task | Function |
|---|---|
| Fit one tree | `TORS(X, y)` |
| Fit a forest | `SWORD(X, y, m = 100)` |
| Predict | `predict(forest, newdata)` |
| OOB convergence | `plot(forest, which = "oob")` |
| OOB scatter | `plot_oob_fit(forest, y)` |
| OIW-VI (variable importance) | `VI_SWORD(forest)` / `plot_vi_sword(forest)` |
| Partial dependence | `pdp_sword(forest, X, "varname")` |
| Tree text print | `plot_tors_text(tree)` |
| Tree graph | `plot(tree)` |
